---
title: "Lieber 2024 TRACE ISA"
author: "Bob Policastro"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, collapse=TRUE, comment="#>")
```

## Libraries

Load the required libraries for the analysis.

```{r, message=FALSE, warning=FALSE}
library("tidyverse")
library("patchwork")
library("randomcoloR")
library("ggseqlogo")
library("Biostrings")
library("karyoploteR")
```

## Prepare Data

Load the annotated insertion sites from the Ensoma TRACE ISA workflow.

```{r}
# Get a list of files to load.
isa_files <-
  file.path("data", "insertion_sites") |>
  list.files(pattern="filtered\\.tsv$", full.names=TRUE)

# Load the files.
isa <- isa_files |>
  tibble(file_path=_) |>
  mutate(
    file_name=basename(file_path),
    data=map(file_path, read_tsv, show_col_types=FALSE)
  ) |>
  unnest(data) |>
  # Remove the Illumina stuff from the sample names.
  mutate(file_id=str_remove(sample_id, "_S\\d+_L00\\d\\.\\S+$")) |>
  relocate(file_id, .after=sample_id)
```

Load the sample sheet that has additional information for each sample.
Then join the information with the ISA results.

```{r}
# Load the sample sheet.
sample_sheet <- read_csv(
  file.path("data", "sample_sheet.csv"), show_col_types=FALSE
)

# Join into the isa data.
isa <- left_join(isa, sample_sheet, by="file_id")
```

Only keep insertion sites that appear in at least 2 of 3 technical
replicates. Return the sum of valies for the group of technical
replicates.

```{r}
isa <- isa |>
  # Keep only insertion sites that appear in at least 2 technical replicates.
  group_by(sample_name, seqnames, start, end, strand) |>
  filter(n() > 1) |>
  ungroup() |>
  # Sum the score for each insertion site across replicates.
  group_by(
    sample_name, seqnames, start, end, strand, annotation, tss_distance 
  ) |>
  summarize(score=sum(score), .groups="drop")
```

Set the factor levels of the sample names so they appear in a consistent
order in the plots.

```{r}
sample_name_order <- c(
  "Mouse PBMC 65.10%", "Mouse PBMC 62.20%", "Mouse PBMC 38.90%",
  "Mouse PBMC 37.00%", "Mouse PBMC 25.00%", "Mouse PBMC 0.00%"
)

isa <- mutate(
  isa, sample_name=factor(sample_name, levels=sample_name_order)
)
```

## Number and Frequency of Insertion Sites

Generate a bar graph of the number of insertion sites per sample. This
is defined as the number of unique genomic locations for each sample
where an insertion event was detected.

```{r, fig.height=3.5, fig.width=4.25}
p <- isa |>
  count(sample_name) |>
  ggplot(aes(x=n, y=fct_rev(sample_name))) +
    geom_col(width=0.75, aes(fill=sample_name), show.legend=FALSE) +
    theme_bw() +
    labs(
      x="Unique Insertion Site\nLocations",
      y="Sample",
      title="Unique Insertion Site\nLocations per Sample"
    ) +
    scale_fill_brewer(palette="Set2")

print(p)

if (!dir.exists("results")) dir.create("results")

ggsave(
  file.path("results", "insertion_sites_per_sample.pdf"), plot=p,
  device=cairo_pdf, height=3.5, width=4.25
)
```

Multiple unique sequenced fragments can end up mapping to the same
genomic location in the same sample. This is likely due to the clonal
expansion of a cell with a particular genomic insertion site. This
bargraph represents all fragments of DNA with insertion sites sequenced
for each sample regardless of genomic location.

```{r, fig.height=3.75, fig.width=4.25}
p <- isa |>
  # Sum the total number of unique fragments for each sample.
  group_by(sample_name) |>
  summarize(score=sum(score), .groups="drop") |>
  # Plot the total number of unique fragments for each sample.
  ggplot(aes(x=score, y=fct_rev(sample_name))) +
    geom_col(width=0.75, aes(fill=sample_name), show.legend=FALSE) +
    theme_bw() +
    theme(axis.text.x=element_text(angle=45, hjust=1)) +
    labs(
      x="Insertion Site-containing\nDNA Fragments Sequenced",
      y="Sample",
      title="Insertion Site-containing\nDNA Fragments Sequenced\nper Sample"
    ) +
    scale_fill_brewer(palette="Set2")

print(p)

ggsave(
  file.path("results", "insertion_sites_sequenced_per_sample.pdf"), plot=p,
  device=cairo_pdf, height=3.5, width=4.25
)
```

As described above, a sample can have multiple DNA fragments with
insertion sites mapping to the same genomic location. This stacked
barplot represents for each genomic location with an insertion site in
the sample, the fraction of insertion-site containing DNA fragments
sequenced associated with that site.

```{r, fig.height=4.5, fig.width=5.5}
# Function to generate random color palette.
random_pal <- function(vals) {
  n_vals <- length(unique(vals))
  color_map <- unique(randomcoloR::randomColor(n_vals + 1000))[1:n_vals]
  names(color_map) <- unique(vals)
  colors <- vapply(vals, \(x) color_map[x], character(1))
  return(colors)
}

# Prepare data for plotting.
df <- isa |>
  # Summarize the fraction of reads per insertion site.
  dplyr::group_by(sample_name) |>
  dplyr::mutate(frac=score / sum(score)) |>
  dplyr::ungroup() |>
  # Add a column with the custom colors.
  dplyr::mutate(
      posid=paste(seqnames, start, end, strand, sep=":"),
      color=random_pal(posid),
      sample_id=factor(
          sample_name,
          levels=rev(stringr::str_sort(unique(sample_name)))
      )
  ) |>
  split(~sample_name) |>
  lapply(\(x) {
      x <- x |>
          dplyr::mutate(
              color=forcats::fct_drop(color),
              color=forcats::fct_reorder(color, frac, .desc=TRUE)
          )
      return(x)
  })

# Create the plots.
p <- df |>
  imap(\(x, y) {
    ggplot(x, aes(x=frac, y=sample_name)) +
      geom_col(aes(fill=color)) +
      scale_fill_identity() +
      scale_x_continuous(labels=scales::percent) +
      theme_classic() +
      labs(
        x="Percentage of Captured\nInsertion Events",
        y="Sample"
      )
  }) |>
  wrap_plots(ncol=1) +
    plot_layout(axes="collect", axis_titles="collect")

print(p)

ggsave(
  file.path("results", "insertion_site_percentage_per_sample.pdf"),
  plot=p, device=cairo_pdf, height=4.5, width=5.5
)
```

## Genomic Distribution

As part of the Ensoma TRACE ISA workflow, each insertion site is
annotated to the nearest genomic feature at the *transcript* level.
Assignment to a feature is based on the the following rules:

-   **Promoter**: -2,000 to +200 bp relative to *transcript*
    transcription start sites (TSS).
-   **Exon**: Non-promoter and non-3\` UTR *exonic* positions.
-   **Intron**: Non-promoter and non-3\` UTR *intronic* positions.
-   **3\` UTR**: From the end of the ORF to the last exon of a
    transcript.
-   **Downstream**: +3,000 kbp from the end of the last exon of the
    transcript.
-   **Distal Intergenic**: Not contained within the above constraints.

```{r, fig.height=4.5, fig.width=5, warning=FALSE}
annotation_factor_order <- c(
  "Promoter", "Exon", "Intron", "3' UTR", "Downstream", "Distal Intergenic"
)

p <- isa |>
  # Remove superfluous parts of the annotation values.
  mutate(annotation=str_remove(annotation, "\\s\\(.+$")) |>
  # Get the fraction of insertion sites associated with each genomic annotation.
  count(sample_name, annotation) |>
  group_by(sample_name) |>
  mutate(frac=n/sum(n)) |>
  ungroup() |>
  complete(sample_name, annotation, fill=list(n=0)) |>
  mutate(annotation=factor(annotation, levels=annotation_factor_order)) |>
  # Create the stacked barplot.
  ggplot(aes(x=frac, y=fct_rev(sample_name))) +
    geom_col(aes(fill=fct_rev(annotation))) +
    scale_fill_brewer(palette="Set1") +
    scale_x_continuous(labels=scales::percent) +
    theme_bw() +
    theme(
      legend.position="bottom"
    ) +
    guides(fill=guide_legend(ncol=2)) +
    labs(
      title="Genomic Distribution of\nInsertion Sites",
      x="Percentage of Insertion Site\nLocations",
      y="Sample",
      fill="Genomic\nAnnotation"
    )

print(p)

ggsave(
  file.path("results", "insertion_site_genomic_locations.pdf"),
  plot=p, device=cairo_pdf, height=4.5, width=5
)
```

The distance to the nearest TSS for each insertion site is calculated as
part of the annotation process. Here we create a histogram of these
distances for each sample, excluding insertion sites greater than +500
kbp or less than -500 kbp from the nearest TSS.

```{r, fig.height=6, fig.width=6}
p <- isa |>
  # Filter out insertion sites where the distance to the nearest TSS
  # is less than -500 kbp or greater than 500 kbp.
  dplyr::filter(abs(tss_distance) < 500000) |>
  # Create the histogram.
  ggplot(aes(x=tss_distance)) +
    geom_histogram(
      bins=100,
      aes(
        y=after_stat(count)/sum(after_stat(count)), fill=sample_name
      ),
      show.legend=FALSE
    ) +
    theme_bw() +
    facet_wrap(
      vars(sample_name), ncol=1, scales="free_y",
      strip.position="right"
    ) +
    scale_x_continuous(
      labels=scales::label_number(suffix=" kbp",scale=1e-3)
    ) +
    scale_y_continuous(labels=scales::percent) +
    scale_fill_brewer(palette="Set2") +
    theme(
      strip.text.y.right=element_text(angle=0),
      strip.background=element_blank(),
      axis.text.y=element_text(size=6)
    ) +
    labs(
      x="Distance to Promoter",
      y="Percent of Insertion Sites",
      title="Distribution of Unique Insertion Site\nLocations Relative to Promoters"
    )

print(p)

ggsave(
  file.path("results", "insertion_sites_promoter_proximity.pdf"),
  plot=p, device=cairo_pdf, height=6, width=6
)
```

For all insertion sites appearing in at least two technical replicates,
create a sequence logo centered around the insertion site including ±20
bps.

Sleeping beauty is known to have a palindromic motif at the insertion
site with A(-3), T(0), A(+1), and T(+4).

```{r, fig.height=8, fig.width=8, warning=FALSE, message=FALSE}
# Load the FASTA data into a table.
fasta_files <- 
  file.path("data", "insertion_site_sequences") |>
  list.files(pattern="\\.fasta$", full.names=TRUE) |>
  tibble(file_path=_) |>
  mutate(
    file_id=str_remove(basename(file_path), "_S\\d+_L00\\d\\.\\S+"),
    data=map(file_path, readDNAStringSet),
    posid=map(data, names),
    seq=map(data, \(x) {unname(as.character(x))})
  ) |>
  left_join(
    select(sample_sheet, file_id, sample_name),
    by="file_id"
  ) |>
  select(!c(data, file_path, file_id)) |>
  unnest(c(posid, seq)) |>
  distinct() |>
  separate_wider_regex(
    posid, c(
      seqnames="^\\S+", ":", start="\\d+", "-", end="\\d+", "\\(",
      strand="[\\+\\-]", "\\)"
    ),
    cols_remove=FALSE
  ) |>
  mutate(across(c(start, end), as.numeric))

# Remove insertion sites that appeared in less than 2 technical replicates.
fasta_files <- semi_join(
  fasta_files,
  mutate(isa, start=start - 21, end=end + 20),
  by=c("sample_name", "seqnames", "start", "end", "strand")
)

# Turn it into a named list for ggseqlogo.
fasta_files <- fasta_files |>
  split(~sample_name) |>
  map(pull, seq)

# Order the sequences.
fasta_files <- fasta_files[levels(isa$sample_name)]

# Create the sequence logos.
p <- ggseqlogo(fasta_files, ncol=1, seq_type="dna") +
  theme(
    text=element_text(size=7)
  ) +
  scale_x_continuous(breaks=seq_len(41), labels=seq(-20, 20, 1)) +
  labs(
    x="Position Relative to\nInsertion Site"
  )

print(p)

ggsave(
  file.path("results", "insertion_site_seqlogo.pdf"),
  plot=p, device=cairo_pdf, height=8, width=8
)
```

For each sample plot the location of the insertion sites relative to
their chromosomal location on a karyogram.

```{r, fig.height=5, fig.width=6}
# Create GRanges for each sample.
is_granges <- isa |>
  select(sample_name, seqnames, start, end, strand, score) |>
  mutate(seqnames=str_c("chr", seqnames)) |>
  split(~ sample_name) |>
  map(makeGRangesFromDataFrame, keep.extra.columns=TRUE)

if (!dir.exists(file.path("results", "karyoplots"))) {
  dir.create(file.path("results", "karyoplots"))
}

# Increase the top plot margin so the title doesn't get cut off.
pp <- getDefaultPlotParams(plot.type=1)
pp$topmargin <- 300

# Create a plot for each sample.
iwalk(is_granges, \(x, y) {
  species <- ifelse(str_starts(y, "Human"), "hg38", "mm39")

  file_name <- y |>
    str_to_lower() |>
    str_replace_all(c(" "="_", "%"="")) |>
    str_glue("{file}_karyoplot.pdf", file=_)

  par(mar=c(1, 1, 10, 1))
  cairo_pdf(
    filename=file.path("results", "karyoplots", file_name),
    height=5, width=6
  )
  
  p <- plotKaryotype(genome=species, main=y, plot.params=pp)
  kpSegments(p, x, y1=0.5, y0=0, col="royalblue")
  
  dev.off()
})
```

## Software Information

```{r, collapse=FALSE, echo=FALSE, comment=""}
sessionInfo()
```
